Casting Multiple Shadows: High-Dimensional Interactive Data Visualisation with Tours and Embeddings

### 2021-06-02 reviews/response.md

Dear editor and reviewers,

We would like to thank for your thorough reviews of our manuscript and
appreciate your constructive criticisms.

In the following response, I would like to address each of your comments
and highlight the amendments we have made to the manuscript and the liminal software.

# Reviewer A

> Unfortunately, I was not able to run the provided code. My guess is that there is some incompatibility with the newest R-version 4-0-3. Please check!

This was a problem with a newer version of the tourr package causing liminal
to not animate the tour correctly. It has been fixed and the package is now 
available on CRAN and has been tested against several versions of R.

# Reviewer B

> For example, the various clusters came pre-colored.  I would like to see a little more about how the analyst would identify these clusters, especially in a case such as your tree example where it doesn't appear that you would have the coloring/clustering you have without a priori knowledge of the different branches.  Is it possible to create that clustering starting from the tour or the embedding?

> Would it be worthwhile to link a pairwise scatterplot matrix for all or a subset of the original variables or components so as to be able to relate clusters and other features via brushing back to those variables for interpretation?  I guess the biplot in Figure 7 may be intended to show something similar?  Is it possible to expand and clarify your results from this figure a bit for those of us who are not genomics specialists?

### 2021-06-03 liminal.Rmd
changes

To begin we suppose the data is in the form of a rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find a low-dimensional representation of the data, 
$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
produces $Y$ will remove noise in the original dataset while retaining any 
latent structure.

@@

The t-SNE algorithm estimates the pair-wise similarity of points in a high
dimensional space based on their (Euclidean) distances using a Gaussian
distribution. A configuration in the low dimensional embedding space is then
estimated by modelling similarities using a t-distribution with 1 degree of
freedom [@Maaten2008-sk]. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

@@

These interactions are enabled by treating the basis generation as a reactive
stream. Instead of realising the entire sequence, which limits the animation
to have a discrete number of frames, new bases and their intermediate frames
are generated dynamically via pushing the current projection to the 
visualisation interface. The interface listens to events like pressing a button
or mouse-dragging and reacts by pausing the stream. This process allows the
user to manipulate the tour in real time rather than having to fix the number
of bases ahead of time. Additionally, once the user has identified an 
interesting projection or is done with the tour, the interface will return the 
current basis for use downstream.

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in Figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (Figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

@@

A common analysis task in single cell studies is performing clustering to
identify groupings  of cells with similar expression profiles. Analysts
in this area generally use non linear DR methods for verification and identification
of clusters and developmental trajectories (i.e., case study 1). For clustering
workflows the primary task is verify the existence of clusters and then begin to 
identify the clusters as cell types using the expression of "known" 
marker genes. Here a 'faithful' a embedding should ideally preserve the topology
of the data; cells that correspond to a cell type should lie in the same
neighbourhood in high-dimensional space. In this case study we use our 
linked brushing approaches to look at neighbourhood preservation
and look at marker genes through the lens of the tour. The data we have
selected for this case study has features similar to those found in case studies
2 and 3.

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed QC using the **scater** package
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using **scran** 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].

>>>>

To begin we suppose the data is in the form of a rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any dimension reduction (DR) algorithm
is to find a low-dimensional representation of the data, 
$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
produces $Y$ will remove noise in the original dataset while retaining any 
latent structure.

@@

The t-SNE algorithm estimates the pair-wise similarity of points in a high
dimensional space based on their (Euclidean) distances using a Gaussian
distribution. A configuration in the low dimensional embedding space is then
estimated by modelling similarities using a t-distribution with 1 degree of
freedom [@Maaten2008-sk]. There are several subtleties of the algorithm that 
are revealed by stepping through its machinery.

@@

These interactions are enabled by treating the basis generation as a reactive
stream. Instead of realising the entire sequence, which limits the animation
to have a discrete number of frames, new bases and their intermediate frames
are generated dynamically via pushing the current projection to the 
visualisation interface. The interface listens to events like pressing a button
or mouse-dragging and reacts by pausing the stream. This process allows the
user to manipulate the tour in real time rather than having to fix the number
of bases ahead of time. Additionally, once the user has identified an 
interesting projection or is done with the tour, the interface will return the 
current basis (as a matrix) for use downstream.

@@

Next we view Gaussian clusters from the 
_Multi Challenge Dataset_, a benchmark simulation data set for clustering tasks
[@Rauber2009-vh]. This dataset has two Gaussian clusters with equal covariance 
embedded in 10-$d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, 
that are each equidistant from each other and have the same covariance structure.
From the video linked in Figure \@ref(fig:hierarchical), we see 
that t-SNE has correctly identified the sub-clusters. However, their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view (Figure \@ref(fig:hierarchical)a). We see 
in this case that it is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent (Figure \@ref(fig:hierarchical)b). When we brush 
the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

@@

A common analysis task in single cell studies is performing clustering to
identify groupings  of cells with similar expression profiles. Analysts
in this area generally use non linear DR methods for verification and identification
of clusters and developmental trajectories (i.e., case study 1). For clustering
workflows the primary task is to verify the existence of clusters and then begin to 
identify the clusters as cell types using the expression of "known" 
marker genes. Here a 'faithful' a embedding should ideally preserve the topology
of the data; cells that correspond to a cell type should lie in the same
neighbourhood in high-dimensional space. In this case study we use our 
linked brushing approaches to look at neighbourhood preservation
and look at marker genes through the lens of the tour. The data we have
selected for this case study has features similar to those found in case studies
2 and 3.

First, we downloaded the raw mouse retinal single cell RNA-seq data from
@Macosko2015-ot using the **scRNAseq** Bioconductor package [@scRNAseq-d]. We
have followed a standard workflow for pre-processing and normalizing this data 
(described by @Amezquita2020-at): we performed quality control using the **scater** package
by removing cells with high proportion of mitochondrial gene expression and low
numbers of genes detected, we log-transformed and normalised the expression 
values and  finally selected highly variable genes (HVGs) using **scran** 
[@McCarthy2017; @Lun2016]. The top ten percent of HVGs were used to subset
the normalised expression matrix and compute PCA using the first 25 components.
Using the PCs we built a shared nearest neighbours graph (with $k = 10$) and 
used Louvain clustering to generate clusters [@Blondel2008-bx].
